Simulating Ultra High-Dimensional Multivariate Data Using the bigsimr R Package

### 2020-10-18 020-simulation-algorithm.Rmd
add cmstats talk notebook and simplify algorithm description

# Simulation algorithm

@@

## Algorithm description

@@

## Simulation Algorithm

@@

### Algorithm 1: Matching Pearson approximately

We aim to simulate multivariate data with a Pearson correlation
matrix ${\bf R}$.

(i) Generate ${\bf X}=(X_1, \ldots, X_d) \sim N_d({\bf 0}, {\bf R})$;  
(ii) Transform ${\bf X}$ to ${\bf U} = (U_1, \ldots,  U_d)$ viz $U_i=\Phi(X_i)$, $i=1, \ldots, d$;  
(iii) Return ${\bf Y}  = (Y_1, \ldots,  Y_d)$, where $Y_i=F_i^{-1}(U_i)$, $i=1, \ldots, d$;

* Where $F_i^{-1}$ is the inverse CDF for the $i^{th}$ gene's marginal
distribution.  
* And ${\bf R}$ is the target Pearson correlation matrix.  
* Pearson's correlation is **not** invariant under the montonically increasing
  transformations in steps 3, 4.  
* So the resulting simulated random vectors will have the **exact** margins but **approximate** ${\bf R}$.

### Algorithm 2: Matching Spearman's exactly

(i) Convert ${\bf R}_{Spearman}$ into ${\bf R}_{Pearson}$ via $\rho_{Pearson} =
  2 \times sin \left( \rho_{Spearman} \times \frac{\pi}{6} \right)$;  
(ii) Generate ${\bf X}=(X_1, \ldots, X_d) \sim N_d({\bf 0}, {\bf} R_{Pearson})$;  
(iii) Transform ${\bf X}$ to ${\bf U} = (U_1, \ldots,  U_d)$ viz $U_i=\Phi(X_i)$, $i=1, \ldots, d$;  
(iv) Return ${\bf Y}  = (Y_1, \ldots,  Y_d)$, where $Y_i=F_i^{-1}(U_i)$, $i=1, \ldots, d$;

- Spearman's $\rho$ relates to the difference in the ranks of pairs of values.
- We take advantage of a closed form correspondence between Pearson and Spearman
  for *normal random variables* in step 1 [@BF17].
- Since Spearman's $\rho$ is invariant under the monotonically increasing
  transformations in steps 3 and 4, the resulting simulated random vectors will
  have the **exact** margins and **exact** Spearman correlation (up to Monte Carlo/numeric error).

### Algorithm 3: Matching Kendall's exactly

(i) Convert ${\bf R}_{Kendall}$ into ${\bf R}_{Pearson}$ via $\rho_{Pearson} = sin \left( \tau \times \frac{\pi}{2} \right)$;  
(ii) Generate ${\bf X}=(X_1, \ldots, X_d) \sim N_d({\bf 0}, {\bf R}_{Pearson})$;  
(iii) Transform ${\bf X}$ to ${\bf U} = (U_1, \ldots,  U_d)$ viz $U_i=\Phi(X_i)$, $i=1, \ldots, d$;  
(iv) Return ${\bf Y}  = (Y_1, \ldots,  Y_d)$, where $Y_i=F_i^{-1}(U_i)$, $i=1, \ldots, d$;

- Kendall's $\tau$ is the difference in the probabilities of concordant and discordant pairs.
- We use the 1-1 relationship between Pearson and Kendall's $\tau$ for *normal
  rvs* in step 1.
- Since Kendall's $\rho$ is invariant under the montonically increasing
  transformations in steps 3 and 4,  
 the resulting simulated random vectors will have the **exact** margins and
 **exact** Kendall's $\tau$  (up to Monte Carlo/numeric error).

One thorny problem when dealing with high-dimensional correlation matrices is
they can become non-positive definite through either during estimation (if done pairwise) or through the algorithm's 
bivariate transformations.

>>>>

# Algorithms

@@

@@

@@

## Random vector generation `bigsimr::rvec` algorithm

<mark>make this algorithm pretty later </mark>

1. Preprocessing for nonparameteric dependency matching.  
	(i) Convert from either ${\bf R_{Spearman}}$ or ${\bf R_{Kendall}}$ into the corresponding MVN input correlation ${\bf R_{Pearson}}$ via \@ref(eq:convertSpearman) or \@ref(eq:convertKendall), respectivity.  
	(ii) Check that ${\bf R_{Pearson}}$ is semi-positive definite.  
	(ii) If not find a close semi-positive ${\bf \widetilde{R}_{Pearson}}$ via `cor_nearPSD()`.  
2. NORTA transformation  
	(1) Generate ${\bf X}=(X_1, \ldots, X_d) \sim N_d({\bf 0}, {\bf R_{Pearson}})$;  
	(2) Transform ${\bf X}$ to ${\bf U} = (U_1, \ldots,  U_d)$ viz $U_i=\Phi(X_i)$, $i=1, \ldots, d$;  
3. Return step  
	(3) Return ${\bf Y}  = (Y_1, \ldots,  Y_d)$, where $Y_i=F_i^{-1}(U_i)$, $i=1, \ldots, d$;

## Other high-performance `bigsimr` algorithms

<mark> ALEX, briefly discuss other algorithms here </mark>

- `cor_bounds()`  
- `cor_covert()` make sure that you are clear this is for only MVN.  
- `cor_fast()`  
- `cor_nearPSD()`
